The low dimensional world of echolocating bats

### 2020-03-01 Paper/main.tex

These limitations of the bat's sonar system suggest that sonar provides the bat with a sparse, low dimensional description of the outside 3D world. Recently, \citet{Geberl2019} confirmed that bio-sonar has a very low spatial resolution. In addition, neurophysiological data suggest that closely spaced objects are not perceived individually by echolocating bats. In contrast, neural responses to echo cascades from densely spaced objects represent a single extended stimulus \citep{Warnecke2018}.

Nevertheless, low dimensional descriptions potentially contain a lot of information and can be sufficient for recognizing locations \citep{Kuc1997b,Kuc1997}. We previously demonstrated that bats might use sonar template-based scene recognition \citep{Vanderelst2016,Vanderelst2017}. In this approach, bats are assumed to compare the acoustic signature of the environment to a set of previously stored signatures. We operationalized these acoustic signatures as time-intensity profiles, which could be derived directly from the cochlear output.

The finding that templates provide sufficient information to recognize locations prompted us to directly evaluate the hypothesis that sonar provides bats inherently with a low dimensional, yet sufficiently rich, description of the world. Testing this, we show that complex synthetic echo sequences can be mapped into a low dimensional space, after filtering them through a model of the bat's auditory periphery. Next, we show that templates derived from real bat habitats can also be projected into this space. Finally, we demonstrate that low dimensional feature space allows recognizing locations in real habitats.

Together these results suggest that echolocating bats live in a low dimensional world. The sonar system and the auditory periphery map the complexity of the 3D environment onto a small number of dimensions. Nevertheless, these low dimensional descriptions are sufficiently rich for bats to recognize locations.

@@

We created artificial echo sequences by randomly spacing ideal point reflectors. The distances of the point reflectors were selected using a two-step process. First we selected $n_{s}$ seed points $s$, uniformly distributed over a range from 0.5 to 7 meters. Next, for every seed point $s$, we selected $n_{c}$ cloud points normally distributed around $s$, with standard deviation 0.25 m. This two step process to modeling natural reflectors was inspired by the results of \citet{Yovel2009}, who found that echoes returning from vegetation could be better modeled using a similar two-step process than simply using uniformly spaced points. We generated xx sets of points for each combination of $n_s \in [xx,xx,xx, xx]$ and $n_c \in [xx, xx, xx, xx]$.

@@

>>>>

We have suggested before that limitations of the bat's sonar system result in bats having access to a sparse, low dimensional description of the outside 3D world only \citep{Vanderelst2015a,Vanderelst2016}. Some recent evidence seems to confirm this. First, \citet{Geberl2019} confirmed that bio-sonar has a very low spatial resolution, both for distance and angular dimensions. In addition, neurophysiological data suggest that closely spaced objects are not perceived individually by echolocating bats. In contrast, neural responses to echo cascades from densely spaced objects seem to represent a single extended stimulus \citep{Warnecke2018}. \todo[inline]{Look at recent Simmons paper where bats refuse to fly through hoop}

Low dimensional descriptions potentially contain a lot of information \citep{Kuc1997b,Kuc1997}. We previously demonstrated that bats might use sonar template-based scene recognition \citep{Vanderelst2016,Vanderelst2017}. In this approach, bats are assumed to compare the acoustic signature of the environment to a set of previously stored signatures. We operationalized these acoustic signatures as time-intensity profiles, which could be derived directly from the cochlear output.

In this paper, we directly evaluate the hypothesis that sonar provides bats inherently with a low dimensional, yet sufficiently rich, description of the world.  We test this hypothesis in two steps. In a first step, we generate a large collection of synthetic echo sequences. We filter these sequences using a model of the bat's auditory periphery \citep{Wiegrebe2008}. The synthetic echo sequences differ greatly in complexity. Nevertheless, we show that their resulting cochlear output can be mapped into a lower dimensional space. In a second step, we show that echoes collected in real bat habitats can be mapped into the same lower dimensional space, without loss of information. We do this by demonstrating that the low dimensional feature space allows recognizing locations in real habitats.

Together these results suggest that echolocating bats live in a low dimensional perceptual world. The physics of the sonar system and the auditory periphery map the complexity of the 3D environment onto a small number of features, irrespective of whether we use artificial or real echo sequences. Nevertheless, these features are sufficiently rich for bats to recognize locations, and as we discuss, support flight control.

@@

We created artificial echo sequences by randomly spacing ideal point reflectors. The distances of the point reflectors were selected using a two-step process. First we selected $n_{s}$ seed points $s$, uniformly distributed over a range from 0.5 to 7 meters. Next, for every seed point $s$, we selected $n_{c}$ cloud points normally distributed around $s$, with standard deviation 0.25 m. This two step process to modeling natural reflectors was inspired by the results of \citet{Yovel2009}, who found that echoes returning from vegetation could be better modeled using a similar two-step process than simply using uniformly spaced points. We generated xx sets of points for each combination of $n_s \in [xx,xx,xx, xx]$ and $n_c \in [xx, xx, xx, xx]$. Synthetic echoes consisting of few reflectors are likely to consist of only weak echoes. To ensure that all sequences included at least one strong echo, we added a single reflector at a distance between xx and xx to each collection of points.

@@

\todo[inline]{Should we talk about the fact that the PCA model fits well across templates generated with different number of reflectors? This seems to depend on the echo having at least one strong (aka nearby) reflector.}
